---
title: "01 Data Preparation"
author: "Tsuyoshi Choh"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This file loads, cleans, and merges all raw data sources required for the dissertation analysis. The output is two cleaned datasets saved as RDS files:

- `cleaned_access_final.rds`: LSOA-level analytical dataset (n = 31,810)
- `lsoa_map.rds`: Spatial dataset with shapefile geometry

---

## 1. Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(sf)
library(stringr)
```

---

## 2. Load Raw Data

### 2.1 LSOA 2011 to 2021 Lookup

```{r load-lookup}
lookup <- read_csv("data/raw/LSOA_2011_2021_lookup.csv")
```

### 2.2 Economic Inactivity (Census 2021, TS066)

```{r load-economic}
economic <- read_csv(
  "data/raw/economic_inactivity_LSOA_2021.csv",
  skip = 7,
  show_col_types = FALSE
)
```

### 2.3 Index of Multiple Deprivation 2019 (File 2)

```{r load-imd}
imd <- read_excel(
  "data/raw/IMD2019_File2_DomainScores.xlsx",
  sheet = 2
)
```

### 2.4 GB Accessibility Indicators (PT)

```{r load-accessibility}
emp_pt <- read_csv("data/raw/accessibility_all/accessibility/employment/access_employment_pt.csv") %>%
  filter(str_starts(geo_code, "E")) %>%
  select(geo_code, employment_45, employment_pct_45)

gp_pt <- read_csv("data/raw/accessibility_all/accessibility/gp/access_gp_pt.csv") %>%
  filter(str_starts(geo_code, "E")) %>%
  select(geo_code, gp_number_45, gp_pct_45)

supermarket_pt <- read_csv("data/raw/accessibility_all/accessibility/supermarket/access_supermarkets_pt.csv") %>%
  filter(str_starts(geo_code, "E")) %>%
  select(geo_code, supermarket_45, supermarket_pct_45)
```

### 2.5 GB Accessibility Indicators (Car)

```{r load-car}
emp_car <- read_csv("data/raw/accessibility_all/accessibility/employment/access_employment_car.csv") %>%
  filter(str_starts(geo_code, "E")) %>%
  select(geo_code, employment_45_car = employment_45)
```

### 2.6 Rural Urban Classification 2021

```{r load-ruc}
ruc <- read_csv("data/raw/RUC_Data.csv") %>%
  select(LSOA21CD, RUC21CD, RUC21NM)
```

### 2.7 LSOA Shapefile

```{r load-shapefile}
lsoa_sf <- st_read("data/raw/Dissertation_shpfile/")
```

---

## 3. Link IMD 2019 to 2021 Boundaries

IMD 2019 uses 2011 LSOA boundaries. We use the ONS lookup to convert to 2021 boundaries. Where one 2011 LSOA maps to multiple 2021 LSOAs, IMD scores are averaged.

```{r imd-link}
imd_2021 <- imd %>%
  left_join(lookup, by = c("LSOA code (2011)" = "LSOA11CD")) %>%
  clean_names() %>%
  group_by(lsoa21cd) %>%
  summarise(
    la_name           = first(lad22nm),
    income_rank       = mean(income_rank_where_1_is_most_deprived, na.rm = TRUE),
    income_decile     = mean(income_decile_where_1_is_most_deprived_10_percent_of_lso_as, na.rm = TRUE),
    education_rank    = mean(education_skills_and_training_rank_where_1_is_most_deprived, na.rm = TRUE),
    education_decile  = mean(education_skills_and_training_decile_where_1_is_most_deprived_10_percent_of_lso_as, na.rm = TRUE),
    employment_rank   = mean(employment_rank_where_1_is_most_deprived, na.rm = TRUE),
    employment_decile = mean(employment_decile_where_1_is_most_deprived_10_percent_of_lso_as, na.rm = TRUE),
    imd_rank          = mean(index_of_multiple_deprivation_imd_rank_where_1_is_most_deprived, na.rm = TRUE),
    imd_decile        = mean(index_of_multiple_deprivation_imd_decile_where_1_is_most_deprived_10_percent_of_lso_as, na.rm = TRUE)
  ) %>%
  ungroup()

nrow(imd_2021)
```


---

## 4. Merge All Datasets

```{r merge}
# Clean economic inactivity data
economic_clean <- economic %>%
  clean_names() %>%
  select(
    lsoa21_code = mnemonic,
    lsoa_name   = area,
    total_16_plus = total_all_usual_residents_aged_16_years_and_over,
    inactive_total = economically_inactive,
    inactive_longterm_sick = economically_inactive_long_term_sick_or_disabled
  ) %>%
  distinct(lsoa21_code, .keep_all = TRUE) %>% 
  mutate(
    inactive_rate      = inactive_total / total_16_plus,
    longterm_sick_rate = inactive_longterm_sick / total_16_plus,
    la_name = lsoa_name %>%
      str_remove("^lsoa2021:") %>%
      str_remove("\\s+\\d+[A-Z]+$")
  )

# Merge all
cleaned_access_final <- economic_clean %>%
  left_join(imd_2021,      by = c("lsoa21_code" = "lsoa21cd")) %>%
  left_join(emp_pt,        by = c("lsoa21_code" = "geo_code")) %>%
  left_join(gp_pt,         by = c("lsoa21_code" = "geo_code")) %>%
  left_join(supermarket_pt,by = c("lsoa21_code" = "geo_code")) %>%
  left_join(emp_car,       by = c("lsoa21_code" = "geo_code")) %>%
  left_join(ruc,           by = c("lsoa21_code" = "LSOA21CD")) %>%
  filter(!is.na(employment_45))

dim(cleaned_access_final)
```

---

## 5. Construct Composite Disadvantage Score

```{r composite-score}
# la_name重複を解消
cleaned_access_final <- cleaned_access_final %>%
  select(-la_name.x) %>%
  rename(la_name = la_name.y)

# Composite score構築
cleaned_access_final <- cleaned_access_final %>%
  mutate(
    income_score      = 1 - rescale(income_decile),
    education_score   = 1 - rescale(education_decile),
    deprivation_score = (income_score + education_score) / 2,
    emp_score         = 1 - rescale(log1p(employment_45)),
    gp_score          = 1 - rescale(log1p(gp_number_45)),
    supermarket_score = 1 - rescale(log1p(supermarket_45)),
    cumulative_score  = (emp_score + gp_score + supermarket_score) / 3,
    composite_score   = (deprivation_score + cumulative_score) / 2,
    pt_car_gap        = employment_45_car - employment_45
  )

summary(cleaned_access_final %>% select(deprivation_score, cumulative_score, composite_score, inactive_rate))
```

---

## 6. Create Spatial Dataset

```{r spatial}
lsoa_map <- lsoa_sf %>%
  filter(str_starts(LSOA21CD, "E")) %>%
  left_join(
    cleaned_access_final %>%
      select(lsoa21_code, la_name, inactive_rate, longterm_sick_rate,
             composite_score, deprivation_score, cumulative_score,
             emp_score, gp_score, supermarket_score,
             employment_45, employment_45_car, pt_car_gap,
             RUC21CD, RUC21NM),
    by = c("LSOA21CD" = "lsoa21_code")
  )

nrow(lsoa_map)
```

---

## 7. Validation Checks

```{r validation}
# Check n
cat("cleaned_access_final rows:", nrow(cleaned_access_final), "\n")
cat("lsoa_map rows:", nrow(lsoa_map), "\n")

# Check NA counts
cat("\nNA counts in key variables:\n")
cleaned_access_final %>%
  select(inactive_rate, composite_score, deprivation_score, cumulative_score) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  print()

# Check score distributions
cat("\nComposite score summary:\n")
summary(cleaned_access_final$composite_score)
```

---

## 8. Save Outputs

```{r save}
saveRDS(cleaned_access_final, "data/processed/cleaned_access_final.rds")
saveRDS(lsoa_map,             "data/processed/lsoa_map.rds")

cat("Saved cleaned_access_final.rds and lsoa_map.rds\n")
```
