---
title: "02 Phase 1: National Analysis"
author: "Tsuyoshi Choh"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This file conducts the national-scale analysis for Phase 1:

1. LA-level ranking by composite disadvantage score
2. Study area selection (South Yorkshire)
3. Global Moran's I (spatial autocorrelation)
4. England-wide maps

---

## 1. Load Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(spdep)
library(ggplot2)
library(scales)

cleaned_access_final <- readRDS("data/processed/cleaned_access_final.rds")
lsoa_map             <- readRDS("data/processed/lsoa_map.rds")

cat("cleaned_access_final:", nrow(cleaned_access_final), "rows\n")
cat("lsoa_map:", nrow(lsoa_map), "rows\n")
```

---

## 2. LA-level Ranking

### 2.1 All LAs

```{r la-ranking-all}
la_ranking_all <- cleaned_access_final %>%
  group_by(la_name) %>%
  summarise(
    total_lsoa        = n(),
    mean_composite    = mean(composite_score,   na.rm = TRUE),
    mean_deprivation  = mean(deprivation_score, na.rm = TRUE),
    mean_accessibility = mean(cumulative_score,  na.rm = TRUE),
    mean_inactivity   = mean(inactive_rate,     na.rm = TRUE)
  ) %>%
  arrange(desc(mean_composite))

head(la_ranking_all, 20)
```

### 2.2 Urban LAs Only (UN1 + UF1, 100+ LSOAs)

```{r la-ranking-urban}
la_ranking_urban <- cleaned_access_final %>%
  filter(RUC21CD %in% c("UN1", "UF1")) %>%
  group_by(la_name) %>%
  summarise(
    total_lsoa        = n(),
    mean_composite    = mean(composite_score,   na.rm = TRUE),
    mean_deprivation  = mean(deprivation_score, na.rm = TRUE),
    mean_accessibility = mean(cumulative_score,  na.rm = TRUE),
    mean_inactivity   = mean(inactive_rate,     na.rm = TRUE)
  ) %>%
  filter(total_lsoa >= 100) %>%
  arrange(desc(mean_composite))

print(la_ranking_urban, n = 20)
```

### 2.3 South Yorkshire Position in Ranking

```{r sy-position}
sy_las <- c("Barnsley", "Doncaster", "Rotherham", "Sheffield")

la_ranking_urban %>%
  mutate(rank = row_number()) %>%
  filter(la_name %in% sy_las) %>%
  select(rank, la_name, total_lsoa, mean_composite, mean_inactivity)
```

---

## 3. Global Moran's I

### 3.1 Build Spatial Weights Matrix

```{r spatial-weights}
lsoa_map_clean <- lsoa_map %>%
  filter(!is.na(composite_score) & !is.na(inactive_rate))

lsoa_sp  <- as(lsoa_map_clean, "Spatial")
neighbors <- poly2nb(lsoa_sp, queen = TRUE)
weights   <- nb2listw(neighbors, style = "W", zero.policy = TRUE)

cat("Spatial weights built for", nrow(lsoa_map_clean), "LSOAs\n")
```

### 3.2 Moran's I: Composite Score

```{r morans-i-composite}
moran_composite <- moran.test(
  as.numeric(lsoa_map_clean$composite_score),
  weights,
  zero.policy = TRUE,
  na.action = na.omit
)

print(moran_composite)
```

### 3.3 Moran's I: Economic Inactivity Rate

```{r morans-i-inactivity}
moran_inactivity <- moran.test(
  as.numeric(lsoa_map_clean$inactive_rate),
  weights,
  zero.policy = TRUE,
  na.action = na.omit
)

print(moran_inactivity)
```

---

## 4. England-wide Maps

### 4.1 Composite Disadvantage Score

```{r map-composite}
ggplot(lsoa_map) +
  geom_sf(aes(fill = composite_score), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Composite\nDisadvantage",
    na.value = "grey80"
  ) +
  labs(
    title = "Composite Disadvantage Score by LSOA",
    subtitle = "England 2021 (n = 31,810)"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white"),
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave("data/processed/map_composite_england.png", width = 10, height = 12, dpi = 300)
```

### 4.2 High Disadvantage LSOAs with South Yorkshire Highlighted

```{r map-high-disadvantage}
# Threshold: top 10% urban LSOAs
threshold_urban <- cleaned_access_final %>%
  filter(RUC21CD %in% c("UN1", "UF1")) %>%
  pull(composite_score) %>%
  quantile(0.9, na.rm = TRUE)

lsoa_map <- lsoa_map %>%
  mutate(
    high_disadvantage  = composite_score >= threshold_urban & RUC21CD %in% c("UN1", "UF1"),
    is_south_yorkshire = str_detect(LSOA21NM, "Barnsley|Doncaster|Rotherham|Sheffield")
  )

# South Yorkshire boundary
south_yorks_boundary <- lsoa_map %>%
  filter(is_south_yorkshire) %>%
  st_union()

ggplot(lsoa_map) +
  geom_sf(fill = "grey90", color = NA) +
  geom_sf(
    data = lsoa_map %>% filter(high_disadvantage == TRUE),
    fill = "#d62728", color = NA, alpha = 0.8
  ) +
  geom_sf(data = south_yorks_boundary, fill = NA, color = "black", linewidth = 0.4) +
  labs(
    title    = "High Disadvantage LSOAs — England",
    subtitle = "Top 10% urban LSOAs by composite score. South Yorkshire outlined."
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white"),
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave("data/processed/map_high_disadvantage_england.png", width = 10, height = 12, dpi = 300)
```

### 4.3 Economic Inactivity Rate

```{r map-inactivity}
ggplot(lsoa_map) +
  geom_sf(aes(fill = inactive_rate), color = NA) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    name = "Inactivity\nRate",
    na.value = "grey80",
    labels = percent_format()
  ) +
  labs(
    title    = "Economic Inactivity Rate by LSOA",
    subtitle = "England 2021 (n = 31,810)"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white"),
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave("data/processed/map_inactivity_england.png", width = 10, height = 12, dpi = 300)
```

---

## 5. Correlation Check

```{r correlation}
cor_overall <- cor(
  cleaned_access_final$composite_score,
  cleaned_access_final$inactive_rate,
  use = "complete.obs"
)

cat("Correlation (composite vs inactivity, England):", round(cor_overall, 3), "\n")

# By urban/rural
cleaned_access_final %>%
  mutate(urban_rural = ifelse(RUC21CD %in% c("UN1", "UF1"), "Urban", "Rural")) %>%
  group_by(urban_rural) %>%
  summarise(
    n           = n(),
    correlation = cor(composite_score, inactive_rate, use = "complete.obs")
  )
```

---

## 6. Save Updated lsoa_map

```{r save}
saveRDS(lsoa_map, "data/processed/lsoa_map.rds")
cat("lsoa_map updated and saved\n")
```
