---
title: "03 Phase 2: South Yorkshire Case Study"
author: "Tsuyoshi Choh"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This file conducts the South Yorkshire case study analysis for Phase 2:

1. South Yorkshire LSOA extraction
2. Comparative analysis (3 disadvantage groups)
3. PT vs Car accessibility gap
4. LISA (local spatial autocorrelation / anomaly detection)
5. Maps

---

## 1. Load Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(spdep)
library(ggplot2)
library(scales)
library(patchwork)

cleaned_access_final <- readRDS("data/processed/cleaned_access_final.rds")
lsoa_map             <- readRDS("data/processed/lsoa_map.rds")
```

---

## 2. Extract South Yorkshire LSOAs

```{r sy-extract}
sy_las <- c("Barnsley", "Doncaster", "Rotherham", "Sheffield")

sy_data <- cleaned_access_final %>%
  filter(la_name %in% sy_las)

sy_map <- lsoa_map %>%
  filter(str_detect(LSOA21NM, "Barnsley|Doncaster|Rotherham|Sheffield"))

cat("South Yorkshire LSOAs:", nrow(sy_data), "\n")
cat("By LA:\n")
sy_data %>% count(la_name) %>% print()
```

---

## 3. Comparative Analysis — 3 Disadvantage Groups

### 3.1 Define Groups

LSOAs are classified into three groups based on the number of accessibility dimensions where they fall in the worst quartile:

- **No disadvantage**: not in worst quartile on any dimension
- **Single disadvantage**: worst quartile on exactly one dimension
- **Cumulative disadvantage**: worst quartile on two or more dimensions

```{r group-classification}
# Thresholds: worst quartile (75th percentile) for each accessibility score
q75_emp   <- quantile(cleaned_access_final$emp_score,         0.75, na.rm = TRUE)
q75_gp    <- quantile(cleaned_access_final$gp_score,          0.75, na.rm = TRUE)
q75_super <- quantile(cleaned_access_final$supermarket_score, 0.75, na.rm = TRUE)

sy_data <- sy_data %>%
  mutate(
    disadv_emp   = emp_score         >= q75_emp,
    disadv_gp    = gp_score          >= q75_gp,
    disadv_super = supermarket_score >= q75_super,
    n_disadvantaged = disadv_emp + disadv_gp + disadv_super,
    disadvantage_group = case_when(
      n_disadvantaged == 0 ~ "No disadvantage",
      n_disadvantaged == 1 ~ "Single disadvantage",
      n_disadvantaged >= 2 ~ "Cumulative disadvantage"
    ),
    disadvantage_group = factor(
      disadvantage_group,
      levels = c("No disadvantage", "Single disadvantage", "Cumulative disadvantage")
    )
  )

sy_data %>% count(disadvantage_group)
```

### 3.2 Compare Inactivity Rates Across Groups

```{r group-comparison}
group_summary <- sy_data %>%
  group_by(disadvantage_group) %>%
  summarise(
    n              = n(),
    mean_inactivity = mean(inactive_rate, na.rm = TRUE),
    median_inactivity = median(inactive_rate, na.rm = TRUE),
    sd_inactivity  = sd(inactive_rate, na.rm = TRUE)
  )

print(group_summary)
```

### 3.3 Visualise

```{r group-boxplot}
ggplot(sy_data, aes(x = disadvantage_group, y = inactive_rate, fill = disadvantage_group)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_fill_manual(values = c("grey70", "#f0a500", "#d62728")) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "Economic Inactivity Rate by Disadvantage Group",
    subtitle = "South Yorkshire LSOAs",
    x        = NULL,
    y        = "Inactivity Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

ggsave("data/processed/sy_group_boxplot.png", width = 9, height = 6, dpi = 300)
```

---

## 4. PT vs Car Accessibility Gap

### 4.1 Summary Statistics

```{r pt-car-summary}
sy_data %>%
  summarise(
    mean_pt    = mean(employment_45,     na.rm = TRUE),
    mean_car   = mean(employment_45_car, na.rm = TRUE),
    mean_gap   = mean(pt_car_gap,        na.rm = TRUE),
    ratio      = mean(employment_45_car, na.rm = TRUE) / mean(employment_45, na.rm = TRUE)
  )
```

### 4.2 Map: PT vs Car Gap

```{r map-pt-car-gap}
sy_map <- lsoa_map %>%
  filter(str_detect(LSOA21NM, "Barnsley|Doncaster|Rotherham|Sheffield")) %>%
  left_join(
    sy_data %>% select(lsoa21_code, disadvantage_group),
    by = c("LSOA21CD" = "lsoa21_code")
  )

ggplot(sy_map) +
  geom_sf(aes(fill = pt_car_gap), color = NA) +
  scale_fill_viridis_c(
    option    = "plasma",
    direction = -1,
    name      = "PT vs Car\nGap",
    na.value  = "grey80"
  ) +
  labs(
    title    = "PT vs Car Employment Accessibility Gap",
    subtitle = "South Yorkshire LSOAs — higher values indicate greater car advantage"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white"),
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave("data/processed/sy_map_pt_car_gap.png", width = 10, height = 8, dpi = 300)
```

---

## 5. LISA — Local Spatial Autocorrelation

### 5.1 Build Spatial Weights

```{r lisa-weights}
sy_map_clean <- sy_map %>%
  filter(!is.na(composite_score) & !is.na(inactive_rate))

sy_sp      <- as(sy_map_clean, "Spatial")
sy_nb      <- poly2nb(sy_sp, queen = TRUE)
sy_weights <- nb2listw(sy_nb, style = "W", zero.policy = TRUE)

cat("LISA weights built for", nrow(sy_map_clean), "LSOAs\n")
```

### 5.2 Local Moran's I: Composite Score

```{r lisa-composite}
lisa_composite <- localmoran(
  as.numeric(sy_map_clean$composite_score),
  sy_weights,
  zero.policy = TRUE,
  na.action   = na.omit
)

sy_map_clean <- sy_map_clean %>%
  mutate(
    lisa_i_composite  = lisa_composite[, "Ii"],
    lisa_p_composite  = lisa_composite[, "Pr(z != E(Ii))"],
    lisa_sig_composite = lisa_p_composite < 0.05
  )
```

### 5.3 Local Moran's I: Inactivity Rate

```{r lisa-inactivity}
lisa_inactivity <- localmoran(
  as.numeric(sy_map_clean$inactive_rate),
  sy_weights,
  zero.policy = TRUE,
  na.action   = na.omit
)

sy_map_clean <- sy_map_clean %>%
  mutate(
    lisa_i_inactivity  = lisa_inactivity[, "Ii"],
    lisa_p_inactivity  = lisa_inactivity[, "Pr(z != E(Ii))"],
    lisa_sig_inactivity = lisa_p_inactivity < 0.05
  )
```

### 5.4 Classify LISA Clusters (HH / HL / LH / LL)

```{r lisa-clusters}
mean_composite  <- mean(sy_map_clean$composite_score, na.rm = TRUE)
mean_inactivity <- mean(sy_map_clean$inactive_rate,   na.rm = TRUE)

sy_map_clean <- sy_map_clean %>%
  mutate(
    # Composite LISA cluster
    lisa_cluster_composite = case_when(
      !lisa_sig_composite                                                        ~ "Not significant",
      composite_score >= mean_composite & lisa_i_composite > 0                  ~ "HH",
      composite_score <  mean_composite & lisa_i_composite > 0                  ~ "LL",
      composite_score >= mean_composite & lisa_i_composite < 0                  ~ "HL",
      composite_score <  mean_composite & lisa_i_composite < 0                  ~ "LH",
      TRUE                                                                       ~ "Not significant"
    ),
    # Inactivity LISA cluster
    lisa_cluster_inactivity = case_when(
      !lisa_sig_inactivity                                                       ~ "Not significant",
      inactive_rate >= mean_inactivity & lisa_i_inactivity > 0                  ~ "HH",
      inactive_rate <  mean_inactivity & lisa_i_inactivity > 0                  ~ "LL",
      inactive_rate >= mean_inactivity & lisa_i_inactivity < 0                  ~ "HL",
      inactive_rate <  mean_inactivity & lisa_i_inactivity < 0                  ~ "LH",
      TRUE                                                                       ~ "Not significant"
    )
  )

cat("LISA clusters (composite):\n")
sy_map_clean %>% st_drop_geometry() %>% count(lisa_cluster_composite) %>% print()

cat("\nLISA clusters (inactivity):\n")
sy_map_clean %>% st_drop_geometry() %>% count(lisa_cluster_inactivity) %>% print()
```

### 5.5 Map LISA Clusters

```{r map-lisa}
cluster_colours <- c(
  "HH"              = "#d62728",
  "LL"              = "#1f77b4",
  "HL"              = "#ff7f0e",
  "LH"              = "#aec7e8",
  "Not significant" = "grey85"
)

map_lisa_composite <- ggplot(sy_map_clean) +
  geom_sf(aes(fill = lisa_cluster_composite), color = NA) +
  scale_fill_manual(values = cluster_colours, name = "Cluster") +
  labs(
    title    = "LISA Clusters: Composite Disadvantage",
    subtitle = "South Yorkshire — p < 0.05"
  ) +
  theme_void(base_size = 12) +
  theme(plot.background = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

map_lisa_inactivity <- ggplot(sy_map_clean) +
  geom_sf(aes(fill = lisa_cluster_inactivity), color = NA) +
  scale_fill_manual(values = cluster_colours, name = "Cluster") +
  labs(
    title    = "LISA Clusters: Economic Inactivity",
    subtitle = "South Yorkshire — p < 0.05"
  ) +
  theme_void(base_size = 12) +
  theme(plot.background = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

map_lisa_composite + map_lisa_inactivity

ggsave("data/processed/sy_lisa_maps.png", width = 14, height = 7, dpi = 300)
```

---

## 6. Anomaly Detection — HL LSOAs

### 6.1 Identify HL LSOAs

```{r hl-anomalies}
hl_lsoas <- sy_map_clean %>%
  st_drop_geometry() %>%
  filter(lisa_cluster_inactivity == "HL") %>%
  select(LSOA21CD, LSOA21NM, inactive_rate, composite_score,
         pt_car_gap, deprivation_score, cumulative_score)

cat("HL LSOAs (inactivity):", nrow(hl_lsoas), "\n")
print(hl_lsoas)
```

### 6.2 Compare HL vs Surrounding LSOAs

```{r hl-comparison}
sy_map_clean %>%
  st_drop_geometry() %>%
  filter(lisa_cluster_inactivity %in% c("HL", "LL", "LH")) %>%
  group_by(lisa_cluster_inactivity) %>%
  summarise(
    n                 = n(),
    mean_inactivity   = mean(inactive_rate,     na.rm = TRUE),
    mean_composite    = mean(composite_score,   na.rm = TRUE),
    mean_pt_car_gap   = mean(pt_car_gap,        na.rm = TRUE),
    mean_deprivation  = mean(deprivation_score, na.rm = TRUE),
    mean_accessibility = mean(cumulative_score,  na.rm = TRUE)
  )
```
```{r}
sy_map_clean %>%
  st_drop_geometry() %>%
  filter(lisa_cluster_inactivity == "HL") %>%
  select(LSOA21NM, inactive_rate, pt_car_gap, composite_score, deprivation_score, cumulative_score) %>%
  arrange(desc(inactive_rate)) %>%
  as.data.frame()
```

```{r}
car_ownership <- read_csv("data/raw/car_ownership_LSOA_2021.csv", skip = 7, show_col_types = FALSE) %>%
  clean_names() %>%
  select(
    lsoa21_code = mnemonic,
    total_households = total_all_households,
    no_car = no_cars_or_vans_in_household
  ) %>%
  mutate(no_car_rate = no_car / total_households)

head(car_ownership)
```

```{r}
sy_map_clean <- sy_map_clean %>%
  left_join(car_ownership %>% select(lsoa21_code, no_car_rate),
            by = c("LSOA21CD" = "lsoa21_code"))

# HL vs LL比較
sy_map_clean %>%
  st_drop_geometry() %>%
  filter(lisa_cluster_inactivity %in% c("HL", "LL", "LH")) %>%
  group_by(lisa_cluster_inactivity) %>%
  summarise(
    n               = n(),
    mean_inactivity = mean(inactive_rate,  na.rm = TRUE),
    mean_pt_car_gap = mean(pt_car_gap,     na.rm = TRUE),
    mean_no_car     = mean(no_car_rate,    na.rm = TRUE),
    mean_composite  = mean(composite_score, na.rm = TRUE)
  )
```

```{r}
sy_data %>%
  summarise(
    cor_emp   = cor(emp_score,         inactive_rate, use = "complete.obs"),
    cor_gp    = cor(gp_score,          inactive_rate, use = "complete.obs"),
    cor_super = cor(supermarket_score, inactive_rate, use = "complete.obs"),
    cor_composite = cor(composite_score, inactive_rate, use = "complete.obs")
  )
```

```{r}
# 現在の閾値を確認
cat("q75_emp:", q75_emp, "\n")
cat("q75_gp:", q75_gp, "\n")
cat("q75_super:", q75_super, "\n")

# England全体の75パーセンタイルと比較
cat("\nEngland全体:\n")
cat("q75_emp_england:", quantile(cleaned_access_final$emp_score, 0.75, na.rm = TRUE), "\n")
cat("q75_gp_england:", quantile(cleaned_access_final$gp_score, 0.75, na.rm = TRUE), "\n")
cat("q75_super_england:", quantile(cleaned_access_final$supermarket_score, 0.75, na.rm = TRUE), "\n")

cor(cleaned_access_final$emp_score,         cleaned_access_final$inactive_rate, use = "complete.obs")
cor(cleaned_access_final$gp_score,          cleaned_access_final$inactive_rate, use = "complete.obs")
cor(cleaned_access_final$supermarket_score, cleaned_access_final$inactive_rate, use = "complete.obs")
cor(cleaned_access_final$cumulative_score,  cleaned_access_final$inactive_rate, use = "complete.obs")
cor(cleaned_access_final$deprivation_score, cleaned_access_final$inactive_rate, use = "complete.obs")

```

```{r}
sy_data <- sy_data %>%
  left_join(car_ownership %>% select(lsoa21_code, no_car_rate),
            by = c("lsoa21_code" = "lsoa21_code"))

cor(sy_data$no_car_rate, sy_data$inactive_rate, use = "complete.obs")
```

```{r}
sy_data %>%
  select(emp_score, gp_score, supermarket_score, cumulative_score, 
         deprivation_score, composite_score, inactive_rate) %>%
  summary()
```

```{r}
# South Yorkshire内のスコアの分布確認
sy_data %>%
  summarise(
    mean_emp   = mean(emp_score, na.rm = TRUE),
    mean_gp    = mean(gp_score, na.rm = TRUE),
    mean_super = mean(supermarket_score, na.rm = TRUE),
    sd_emp     = sd(emp_score, na.rm = TRUE),
    sd_gp      = sd(gp_score, na.rm = TRUE),
    sd_super   = sd(supermarket_score, na.rm = TRUE)
  )
```

```{r}
sy_data %>%
  summarise(
    cor_emp        = cor(emp_score,         inactive_rate, use = "complete.obs"),
    cor_gp         = cor(gp_score,          inactive_rate, use = "complete.obs"),
    cor_super      = cor(supermarket_score, inactive_rate, use = "complete.obs"),
    cor_cumulative = cor(cumulative_score,  inactive_rate, use = "complete.obs"),
    cor_deprivation = cor(deprivation_score, inactive_rate, use = "complete.obs"),
    cor_composite  = cor(composite_score,   inactive_rate, use = "complete.obs")
  )
```



---

## 7. Save Outputs

```{r save}
saveRDS(sy_data,     "data/processed/sy_data.rds")
saveRDS(sy_map_clean,"data/processed/sy_map_clean.rds")

cat("Saved sy_data.rds and sy_map_clean.rds\n")
```
